Data Analysis Methods for the Detection of the Epoch of Reionization

### 2017-11-22 PAPER.tex
made jonnie changes until section 3.3

In a $21$ cm power spectrum, detections could be the EoR signal, but they could also (and unfortunately more likely) be attributed to other sources of bias. Connecting a detection to EoR as opposed to noise or foreground bias is a key challenge of future $21$ cm data analyses. In this section we will discuss possible sources of bias in a measurement, as well as techniques that can help mitigate their effects. We will also present a series of tests which we suggest be used to help evaluate deep limits and/or detections.

@@

We now approach the difficult task of tracing excesses to foreground, noise, and EoR biases through a discussion of useful jackknife tests. The jackknife is a resampling technique in which a statistic (i.e. power spectrum) is computed in subsets of the data. These subsets are then compared to reveal systematics. In this section we define two main tests --- the null test and the traditional jackknife --- and explain how a power spectrum detection must pass each. We then highlight how these tests can be used to help distinguish between different sources of bias.

@@

The sensitivity prediction (\citealt{parsons_et_al2012a}) for a power spectral analysis of interferometric $21$ cm data, in temperature-units, is:

@@

where $\nu$ are frequencies in GHz. We use a receiver temperature of $200$ K, yielding $T_{sys} = 487$ K at $150$ MHz. This is lower than in P64 because of previous failure to correct for using a median statistic.
\item $\sqrt{2}$: This factor in the denominator of the sensitivity equation comes from taking the real part of the power spectrum estimates after squaring visibility measurements. In P64, a factor of $2$ was mistakenly used.
\item $N_{lsts}$: The number of LST hours that go into a power spectrum estimation. The sensitivity scales as the square root because we integrate incoherently over time. For PAPER-64, $N_{lsts} = 8$ hours.
\item $N_{seps}$: The number of baseline separation types averaged incoherently in a final power spectrum estimate. For the analysis in this paper, we only use one type of baseline, hence $N_{seps}=1$. The updated limits in \cc{cite kolopanis paper} use three separation types.
\item $t_{int}$: The integration time of the data. It is crucial to adapt this number if filtering is applied along the time axis (i.e. a fringe-rate filter). We compute the effective integration time of our fringe-rate filtered data by scaling the original integration time using the following:
\begin{equation}
t_{frf} = t_{int} \frac{\int1 \, df}{\int w^{2}(f) \,df},
\end{equation}
where $t_{int}=43$ seconds, $t_{frf}$ is the fringe-rate filtered integration time, $w$ is the fringe-rate profile, and the integral is taken over all fringe-rates. For PAPER-64, this number is $t_{int} = 3857$ s. 
\item $N_{days}$: The total number of days of data analyzed. In P64, this number was set to $135$. However, because we divide our data in half (to form `even' and `odd' datasets), this number should be reduced by a factor of $2$. Additionally, because our LST coverage is not $100\%$ complete (it doesn't overlap for every single day), we compute a realistic value of $N_{days}$ as:
\begin{equation}
 \frac{1}{N_{days}} = \sqrt{\Big\langle\frac{1}{N_{i}^{2}} \Big\rangle_{i}},
 \end{equation}
\noindent where $i$ is over LST (\citealt{jacobs_et_al2015}). For PAPER-64, our revised estimate of $N_{days}$ is $\sim34$ days.
\item $N_{bls}$: The number of baselines contributing to the sensitivity of a power spectrum estimate. In P64, this number was the total number of $30$ m East/West baselines used in the analysis. However, using the total number of baselines neglects the fact that baselines are averaged into groups before cross-multiplying data. Our revised estimate for the parameter is:
\begin{equation}
N_{bls} = \frac{N_{bls}}{N_{gps}}\sqrt{N_{gps}^{2}-N_{gps}},
\end{equation}
\noindent where, for the P64 analysis, $N_{gps} = 5$. Each baseline group averages down linearly as the number of baselines entering the group ($N_{bls}/N_{gps}$) and then as the square root of the number of cross-multiplied pairs ($\sqrt{N_{gps}^{2} - N_{gps}}$). For the revised PAPER-64 analysis with only one baseline separation type, this becomes $N_{bls} \sim 46$. 
\item $N_{pols}$: The number of polarizations averaged together. For the case of Stokes I, $N_{pols}=2$.
\end{itemize}

@@

Power spectrum results for the noise simulation are shown in Figure \ref{fig:ps_noise}, where the black and grey points represent positive and negative power spectrum values, respectively (with $2\sigma$ error bars and weighting matrix $\hat{\textbf{C}}_{eff}$), the dashed blue line represents the unweighted power spectrum, and the solid green line denotes our $2\sigma$ theoretical noise prediction as calculated by Equation \eqref{eq:sense}. All three are in agreement, validating our analytical thermal noise calculation.

@@

As discussed in Section \ref{sec:ErrorOverview}, bootstrapping is only a valid way of estimating errors if a dataset is comprised of independent samples. The PAPER-64 pipeline outputs $20$ bootstraps (over baselines), each a $2$-dimensional power spectrum that is a function of $k$ and time. In P64, a second round of bootstrapping occurred over the baseline bootstrap and time axes simultaneously. Random values were sampled with replacement along both axes, drawing as many values as there are number of time integrations. Final power spectrum limits were then computed by taking the mean and standard deviation over this second bootstrap axis (which is typically large, on the order of $N_{boot} \sim 400$). We emphasize that in this previous version, the number of elements sampled per bootstrap ($N \sim $ number of times $\sim$ $700$) greatly exceeds the number of independent LST samples ($\sim$$8$), under-estimating errors.

Because our dataset is fringe-rate filtered, we have substantially increased our integration time, resulting in fewer independent samples (from $\sim700$ samples to $8$ samples). A random draw of $700$ measurements from this dataset therefore has many repeated values, and the variance between hundreds ($N_{boot}$) of these random samples is smaller than the true underlying variance of the data.

Given our new understanding of the sensitivity of bootstraps to the number of elements sampled, we have removed the second step entirely and now simply bootstrap over baselines with the ``single replacement" strategy described in Section \ref{sec:ErrorOverview}.

@@

Power spectrum estimates with these bootstrapping changes for PAPER-64 fringe-rate filtered data are shown in Figure \ref{fig:data_errors}. The estimates are unweighted in order to disentangle the effects of bootstrapping from signal loss. As shown in the figure, when the baseline average includes many repeated samples the error is over-estimated, while when only one sample in the same average is allowed the possibility of being a repeat, the error on the noise-dominated higher $k$ modes converges to the theoretical noise level.  However, if more elements are drawn for each bootstrap than the number of independent samples (by over-sampling elements along the time axis), repeated values begin to crop up and the apparent variation between bootstraps drops, resulting in limits below the predicted noise level.

>>>>

In a $21$ cm power spectrum, detections could be the EoR signal, but they could also (and unfortunately more likely) be attributed to other sources of bias. Connecting a detection to EoR as opposed to noise or foreground bias is a key challenge of future $21$ cm data analyses. In this section we will discuss possible sources of bias in a measurement, as well as techniques that can help mitigate their effects. We will also present a series of tests in a pedagogical fashion which we suggest be used to help evaluate deep limits and/or detections.

@@

We now approach the difficult task of tracing excesses to foreground, noise, and EoR biases through a discussion of useful jackknife tests. Again, we first approach this topic pedagogically as an introduction to the related PAPER discussion in Section \ref{sec:Bias}.

The jackknife is a resampling technique in which a statistic (i.e. power spectrum) is computed in subsets of the data. These subsets are then compared to reveal systematics. In this section we define two main tests --- the null test and the traditional jackknife --- and explain how a power spectrum detection must pass each. We then highlight how these tests can be used to help distinguish between different sources of bias.

@@

The sensitivity prediction (\citealt{parsons_et_al2012a}, \citealt{pober_et_al2013}) for a power spectral analysis of interferometric $21$ cm data, in temperature-units, is:

@@

where $\nu$ are frequencies in GHz. We use a receiver temperature of $200$ K, yielding $T_{sys} = 487$ K at $150$ MHz. This is lower than in P64 because of previous failure to correct for using a median statistic.
\item $\sqrt{2}$: This factor in the denominator of the sensitivity equation comes from taking the real part of the power spectrum estimates after squaring visibility measurements. In P64, a factor of $2$ was mistakenly used.
\item $N_{lsts}$: The number of LST hours that go into a power spectrum estimation. The sensitivity scales as the square root because we integrate incoherently over time. For PAPER-64, $N_{lsts} = 8$ hours.
\item $N_{seps}$: The number of baseline separation types averaged incoherently in a final power spectrum estimate. For the analysis in this paper, we only use one type of baseline, hence $N_{seps}=1$. The updated limits in \cc{cite kolopanis paper} use three separation types.
\item $t_{int}$: The integration time of the data. It is crucial to adapt this number if filtering is applied along the time axis (i.e. a fringe-rate filter). We compute the effective integration time of our fringe-rate filtered data by scaling the original integration time using the following:
\begin{equation}
t_{frf} = t_{int} \frac{\int1 \, df}{\int w^{2}(f) \,df},
\end{equation}
where $t_{int}=43$ seconds, $t_{frf}$ is the fringe-rate filtered integration time, $w$ is the fringe-rate profile, and the integral is taken over all fringe-rates. For PAPER-64, this number is $t_{int} = 3857$ s. 
\item $N_{days}$: The total number of days of data analyzed. In P64, this number was set to $135$. However, because we divide our data in half (to form `even' and `odd' datasets), this number should be reduced by a factor of $2$. Additionally, because our LST coverage is not $100\%$ complete (it doesn't overlap for every single day), we compute a realistic value of $N_{days}$ as:
\begin{equation}
 \frac{1}{N_{days}} = \sqrt{\Big\langle\frac{1}{N_{i}^{2}} \Big\rangle_{i}},
 \end{equation}
\noindent where $i$ is over LST (\citealt{jacobs_et_al2015}). For PAPER-64, our revised estimate of $N_{days}$ is $\sim34$ days.
\item $N_{bls}$: The number of baselines contributing to the sensitivity of a power spectrum estimate. In P64, this number was the total number of $30$ m East/West baselines used in the analysis. However, using the total number of baselines neglects the fact that baselines are averaged into groups before cross-multiplying data. Our revised estimate for the parameter is:
\begin{equation}
N_{bls} = \frac{N_{bls}}{N_{gps}}\sqrt{N_{gps}^{2}-N_{gps}},
\end{equation}
\noindent where, for the P64 analysis, $N_{gps} = 5$. Each baseline group averages down linearly as the number of baselines entering the group ($N_{bls}/N_{gps}$) and then as the square root of the number of cross-multiplied pairs ($\sqrt{N_{gps}^{2} - N_{gps}}$). For the revised PAPER-64 analysis with only one baseline separation type, this becomes $N_{bls} \sim 46$ instead of $51$. 
\item $N_{pols}$: The number of polarizations averaged together. For the case of Stokes I, $N_{pols}=2$.
\end{itemize}

@@

Power spectrum results for the noise simulation, which uses our full power spectrum pipeline, are shown in Figure \ref{fig:ps_noise}, where the black and grey points represent positive and negative power spectrum values, respectively (with $2\sigma$ error bars and weighting matrix $\hat{\textbf{C}}_{eff}$), the dashed blue line represents the unweighted power spectrum, and the solid green line denotes our $2\sigma$ theoretical noise prediction as calculated by Equation \eqref{eq:sense}. All three are in agreement, validating our analytical thermal noise calculation.

@@

As discussed in Section \ref{sec:ErrorOverview}, bootstrapping is only a valid way of estimating errors if a dataset is comprised of independent samples. The PAPER-64 pipeline outputs $20$ bootstraps (over baselines), each a $2$-dimensional power spectrum that is a function of $k$ and time.

In P64, a second round of bootstrapping occurred over the time axis. A total of $400$ bootstraps were created in this step ($N_{boot} = 400$), each comprised of randomly selected values sampled with replacement along the time axis. More specifically, each of these bootstraps contained the same number of values as the number of time integrations (which, at $\sim$$700$, greatly exceeds the approximate number of independent samples after fringe-rate filtering, which is $\sim$$8$). Medians were then taken of the values in each bootstrap (with the appropriate median correction factor applied). Finally, power spectrum limits were computed by taking the mean and standard deviation over all the bootstraps. We emphasize again that in this previous analysis, the number of elements sampled per bootstrap ($N \sim $ number of times $\sim$ $700$) greatly exceeded the number of independent LST samples ($\sim$$8$), under-estimating errors. A random draw of $700$ measurements from this dataset has many repeated values, and the variance between hundreds ($N_{boot}$) of these random samples is smaller than the true underlying variance of the data.

Given our new understanding of the sensitivity of bootstraps to the number of elements sampled, we have removed the second bootstrapping step along time entirely and now simply bootstrap over baselines with the ``single replacement" strategy described in Section \ref{sec:ErrorOverview}.

@@

Power spectrum estimates with these bootstrapping changes for PAPER-64 fringe-rate filtered data are shown in Figure \ref{fig:data_errors}. The estimates are unweighted in order to disentangle the effects of bootstrapping from signal loss. As shown in the figure, when the baseline average includes many repeated samples our sensitivity is not maximized, while when only one sample in the same average is allowed the possibility of being a repeat, the error on the noise-dominated higher $k$ modes converges to the theoretical noise level.  However, if more elements are drawn for each bootstrap than the number of independent samples (by over-sampling elements along the time axis), repeated values begin to crop up and the apparent variation between bootstraps drops, resulting in limits below the predicted noise level.
